Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-05-14 analysis/paper/article.qmd
working on JB's comments

Junior scholars with faculty ambitions are thus stuck in a double-bind, forced to funnel their energy into seemingly endless applications for which they receive little to no feedback and which remain invisible on their CVs. As Dennis et al. underscore, early career researchers are "supplicants"—"There is no way for applicants to point out that the job ads are taking too much time out of the scholarly community’s collective time bank" [@dennisWorstAnthroJob2022, pp. 1].Acknowledging the new reality that applying for academic positions is in and of itself a full-time job, a growing number of anthropology graduate programmes are offering professionalization courses that cover the ins and outs of everything from crafting strong cover letters and CVs to performing well in long-list interviews and preparing teaching demonstrations. These courses can provide valuable training in the hidden curriculum of the academy, but the fact that job applications are so complicated that they require training to navigate suggests more disciplinary attention should be paid to the "market" as a historically contingent set of cultural practices. As Rabinow maintains, "if we want ethical considerations to play a central role in the articulation of truth and power—and I think we do—then bringing such considerations into view is the necessary first step towards recognizing who we are today and setting out on the road to a better place" [-@rabinow1992, pp. 71].

@@

A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, and inclusion. This requirement has been much-debated largely due to an experiment during 2016–2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about the merits and risks of requiring and using diversity statements. These debates drew attention to the challenges of hiring faculty from underrepresented minorities, and resulted in a variety of approaches to evaluating diversity, even leading some universities to entirely omit requirements for diversity statements in job applications [@guidenAreDiversityStatements]. **MORE DETAILS ON THIS HERE—JB WILL ADD.**

# Methods

<!--JB: Given that we have expanded our section on the wiki by several paragraphs, it might be useful for us to divide the methods into two subsections: (1) Dynamics of the wiki, and (2) Scraping the wiki and categorizing ads. Feel free to pick better titles, but I think this would provide more guidance for our readers so they do not get lost in the welter of details about the wiki. Lmk what you think -->

Our primary data source is the Archaeology Academic Jobs Wiki. Originating in 2007, the wiki is a set of freely accessible web pages that anyone can edit anonymously or with a free user account. The site is hosted by Fandom, a for-profit company. The Archaeology pages are part of the Academic Jobs Wiki, which coordinates similar collaboratively-edited resources for around 40 academic disciplines. The coordinators and contributors are nearly all anonymous or pseudonymous. Typically, contributors copy and paste all or some of the text of job ads into the wiki, from a variety of sources such as the *Chronicle of Higher Education*, *Higher Ed Jobs*, and university websites. Other contributors then edit the web page to add comments below an ad to share relevant information based on their experience in applying for that position. These edits result in annotations such as a tally of how many people have applied, the dates of events such as requests for more materials, interviews, offer made, rejection notices, etc. Contributors also edit the page to ask and answer questions about the positions and the application process. These comments have made <!--JB: If we are a bit more slippery about our tenses, we don't need to argue that the wiki is STILL crucial, esp. if late we will discuss its decline--> the Academic Jobs Wiki a unique resource for timely and specific information for job-seekers about positions they are interested in, and one of the most important internet resources for the academic job market. Because of its reputation for aggregating ads from diverse sources and rapidly-updated information that is not available elsewhere, the Academic Jobs Wiki has become an authoritative data source for studies of hiring trends in academia [e.g. @Passalacqua_2018; @musial2018five] and a widely recommended resource for applicants [e.g. @lightfoot2021preparing].

@@

Collaboratively-edited resources such as the Academic Jobs Wiki can be highly variable in the amount and type of editorial activity over time. It is important to characterise this activity to assess the reliability of the content is as a source of information about the job market. The MediaWiki technology used by Academic Jobs Wiki creates a public record of every act of adding or removing text to any page in the Wiki [@barrett2008mediawiki]. @fig-panel-wiki-characteristics, which allows us to assess the dynamics of editing activity over the ten years of our study period. Every edit includes the author's identity, recorded as either an IP address (a numerical address that includes information about the location of the user's computer) <!--good call being explicit here--> or a username. Of the `r n_all_user_ids` unique editors in our sample, `r per_users_we_can_geolocate`% used IP addresses rather than usernames, and geolocation analysis shows that `r results_ip_addresses_tbl_us_scl`% of IP addresses were based in the US, indicating the Archaeology Jobs Wiki is primarily edited and visited by users in the U.S. <!--JB: technically we don't know if they are archaeologists; could be people from other disciplines e.g. Middle Eastern Studies, history etc.-->. The number of edits and active editors was roughly halved from 2013 to 2017, and then continued to gradually decline. Despite this downward trend, the size of individual edits slightly increases over time, and the number of edits per job ad is relatively constant across our study period. The majority of editors are active on the site for less than a year (n = r`edits_for_each_year_tbl_span_tbl_less_than_one_year`, `r edits_for_each_year_tbl_span_tbl_less_than_one_year_perc`%), perhaps reflecting a single season of job-searching. Editors with a longer history likely come from two groups, job seekers participating in multiple seasons of job applications, and faculty on search committees posting multiple job ads.

<!--JB: Ben, this is where I'd like some of the details about min, max, etc. What proportion of people are on the wiki for more than a year, and how long do they stay on? ALSO, and I just thought of this, if ECRs are highly mobile, the same ECR might be represented by different, year-long IP addresses over their period of job seeking. For example, when I was on the market, I lived in Michigan, Pennsylvania, New York, England, and Massachusetts—would I not look like five distinct users who each spent 1-2 years on the wiki using our rationale here? This perhaps merits further discussion) -->

@@

@@

The entire R code [@rlanguage] and data files used for all the analyses and visualizations contained in this paper are openly available at https://doi.org/xxx/xx \[ for peer reviewers the link is https://zenodo.org/records/14798942?preview=1&token=eyJhbGciOiJIUzUxMiJ9.eyJpZCI6ImMyYWE2MDc3LTMxMzAtNDM1Yi1hYTJkLWMyMGMyNWVhOWMyYiIsImRhdGEiOnt9LCJyYW5kb20iOiJiMjkyMTAyN2UwNzEwNWQ2OWIyZmQ4OTkwMjU0MzY1YyJ9.slnC1BbNgZkiJjrRqmKFK9a1bpI-e9VBbjcU7L5OEeYFs3whcL3POsfVj7phDnoaB5A_6mP9wpGirEaT9Dtvug \] to enable re-use of materials and improve reproducibility and transparency [@Marwick2017]. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and the figures as CC-BY, to enable maximum re-use.

@@

Panel A of @fig-show-map-of-hiring-institution-and-cc shows the frequencies of <!-- should we specify "hiring" institutions here?--> institutions according to their Carnegie Classification, a framework for classifying US colleges and universities according to the types of degrees awarded, levels of research activity, and topical foci [@shulman2001carnegie]. Doctoral universities with high and very high research activity are by far the most active institutions hiring archaeology faculty. Associate's colleges, also known as community colleges, rarely post job ads for archaeology faculty.

@@

Demand for jobs focusing on the Americas has been generally high over the past decade, with a peak in 2019-2020 and a subsequent decrease. Demand for jobs focusing on Africa was very low until 2019-2020, peaking in 2020-2021. The proportion of ads with a geographic focus on the Mediterranean has varied substantially, peaking in 2016 and experiencing a nadir in 2019, showing an inverse pattern to the Americas. Asia and India, the Near East, and Europe <!--crying face emoji--> rarely occur as a geographical focus in job ads at US institutions. Asia and India, Africa, and the Americas appear correlated with each other, while the Near East and Mediterranean are inversely correlated in an opposite trend.

@@

Landscape archaeology, encompassing GIS and remote sensing, has remained prominent compared to other methods [@fig-show-metho-trends]. The popularity of this suite of skills may reflect its demand by employers in the Cultural Research Management (CRM) sector. @morganReadyNotArchaeological2023 found that almost one-third of 599 jobs ads posted by CRM employers sought candidates with experience in GIS <!--JB: We're using "methods" a lot in this paragraph, so to avoid redundancy I would recommend writing GIS skills or simply GIS-->. Methods focused on a specific element of the archaeological record, such as Lithic analysis, Zooarchaeology and Ceramics are among the least frequently mentioned in job ads. Instead, more popular methods are those that are relevant to multiple elements of the archaeological record (e.g. Archaeobotany encompasses macroscopic and microscopic plant remains; Bioarchaeology often includes skeletal analysis, isotopes, proteins, etc.).

@@

This discrepancy in our data set—wherein advertisements for tenure-track positions dominate the wiki despite their increasing rarity in larger market—may be due to several factors. The first involves the more limited circulation of advertising for short-term positions relative to advertising for tenure-track jobs. Many of these short-term positions are not advertised nationally, but are instead disseminated through local email lists and are filled by people close to the hiring department, such as recently graduated students. A second factor is bias in our data—with most Wiki users likely seeking a tenure-track job, non-tenure-track jobs may have been less frequently added to the Academic Jobs Wiki because they were peripheral to the goal of most users. This bias limits the reliability of our results on the ratio of tenure-track to non-tenure-track positions. Tellingly, Passalacqua’s examination of a similar wiki over a seven-year span showed a similar dominance of permanent positions—291/474 (63%) were for tenure-track hires at various levels (2018: Table 2). This suggests that these patterns are a reflection of the wiki, rather than the market itself. <!--JB MAYBE ADD A NOTE HERE FROM A CRISIS PAPER, PERHAPS MELISSA'S, ABOUT DISPARITY IN TRAINING AND APPLIED ANTHROPOLOGY. -->

@@

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. Over the course of the study period, we document a gradual decline in topics such as the Archaeology of complex societies and Archaeological science, the geographic foci of Mesoamerica and South America, and methods relating to Landscape archaeology <!--There is some shifting throughout the paper as to whether we capitalize the first letter of region/methods/topics or not. I think we should be consistent and capitalize throughout-->. These trends are harder to explain as we cannot link their origins to a historical event such as the COVID-19 pandemic. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, relational perspectives, in combination with human subjectivity and cultural framing, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism [@hodderArchaeologicalProcessIntroduction1999; @johnson2019archaeological; @Fogelin2019] <!--Should we cite Cate Frieman here?-->. This theoretical trend might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical laboratory methods for technological, functional, and compositional analyses. Other factors relevant to this decline may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past and an increased concern with climate change, environmental sustainability, and resilience, shifting attention away from the study of monumental architecture, elite societies, political hierarchies, and state systems.

@@

Thanks to the anonymous contributors to the Academic Jobs Wiki, without whose efforts to collect and organise hundreds of job ads, this paper would not exist. To those of you who contributed to the wiki but did not obtain a tenure-track job, and/or decided to pivot to another industry, we hope this paper shows your time on the wiki wa not wasted. Your efforts have helped contribute to more accurate expectation-setting for future generations of archaeologists.

>>>>

Junior scholars with faculty ambitions are thus stuck in a double-bind, forced to funnel their energy into seemingly endless applications for which they receive little to no feedback and which remain invisible on their CVs. As Dennis et al. underscore, early career researchers are "supplicants"—"There is no way for applicants to point out that the job ads are taking too much time out of the scholarly community’s collective time bank" [@dennisWorstAnthroJob2022, pp. 1]. Acknowledging the new reality that applying for academic positions is in and of itself a full-time job, a growing number of anthropology graduate programmes are offering professionalization courses that cover the ins and outs of everything from crafting strong cover letters and CVs to performing well in long-list interviews and preparing teaching demonstrations. These courses can provide valuable training in the hidden curriculum of the academy, but the fact that job applications are so complicated that they require training to navigate suggests more disciplinary attention should be paid to the "market" as a historically contingent set of cultural practices. As Rabinow maintains, "if we want ethical considerations to play a central role in the articulation of truth and power—and I think we do—then bringing such considerations into view is the necessary first step towards recognizing who we are today and setting out on the road to a better place" [-@rabinow1992, pp. 71].

@@

A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, and inclusion. This requirement has been much-debated largely due to an experiment during 2016–2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about the merits and risks of requiring and using diversity statements. These debates drew attention to the challenges of hiring faculty from underrepresented minorities, and resulted in a variety of approaches to evaluating diversity, even leading some universities to entirely omit requirements for diversity statements in job applications [@guidenAreDiversityStatements]. <!-- **MORE DETAILS ON THIS HERE—JB WILL ADD.** BM: ok, thank you! -->

# Materials and methods

## Contextualising the Archaeology Academic Jobs Wiki as a data source

Our primary data source is the Archaeology Academic Jobs Wiki. Originating in 2007, the wiki is a set of freely accessible web pages that anyone can edit anonymously or with a free user account. The site is hosted by Fandom, a for-profit company. The Archaeology pages are part of the Academic Jobs Wiki, which coordinates similar collaboratively-edited resources for around 40 academic disciplines. The coordinators and contributors are nearly all anonymous or pseudonymous. Typically, contributors copy and paste all or some of the text of job ads into the wiki, from a variety of sources such as the *Chronicle of Higher Education*, *Higher Ed Jobs*, and university websites. Other contributors then edit the web page to add comments below an ad to share relevant information based on their experience in applying for that position. These edits result in annotations such as a tally of how many people have applied, the dates of events such as requests for more materials, interviews, offer made, rejection notices, etc. Contributors also edit the page to ask and answer questions about the positions and the application process. These comments make the Academic Jobs Wiki a unique resource for timely and specific information for job-seekers about positions they are interested in, and one of the most important internet resources for the academic job market. Because of its reputation for aggregating ads from diverse sources and rapidly-updated information that is not available elsewhere, the Academic Jobs Wiki has become an authoritative data source for studies of hiring trends in academia [e.g. @Passalacqua_2018; @musial2018five] and a widely recommended resource for applicants [e.g. @lightfoot2021preparing].

@@

Collaboratively-edited resources such as the Academic Jobs Wiki can be highly variable in the amount and type of editorial activity over time. It is important to characterise this activity to assess the reliability of the content is as a source of information about the job market. The MediaWiki technology used by Academic Jobs Wiki creates a public record of every act of adding or removing text to any page in the Wiki [@barrett2008mediawiki]. @fig-panel-wiki-characteristics, which allows us to assess the dynamics of editing activity over the ten years of our study period. Every edit includes the author's identity, recorded as either an IP address (a numerical address that includes information about the location of the user's computer) or a username. Of the `r n_all_user_ids` unique editors in our sample, `r per_users_we_can_geolocate`% used IP addresses rather than usernames, and geolocation analysis shows that `r results_ip_addresses_tbl_us_scl`% of IP addresses were based in the US, indicating the Archaeology Jobs Wiki is primarily edited and visited by users in the U.S. The number of edits and active editors roughly halved from 2013 to 2017, and then continued to gradually decline. Despite this downward trend, the size of individual edits slightly increases over time, and the number of edits per job ad is relatively constant across our study period. The majority of editors are active on the site for less than a year (n = r`edits_for_each_year_tbl_span_tbl_less_than_one_year`, `r edits_for_each_year_tbl_span_tbl_less_than_one_year_perc`%), perhaps reflecting a single season of job-searching. Editors with a longer history likely come from two groups, job seekers participating in multiple seasons of job applications, and faculty on search committees posting multiple job ads. That said, inferring the number of editors from the number of IP addresses is complicated by the high mobility of job seekers. For example one person could be represented by several IP addresses as they relocate from one city to another while working in a series of short-term appointments such as post-doctoral fellowships and adjunct teaching positions. This might result in an overestimation of the number of editors on the Wiki and underestimation of the duration of their editorial activity.

<!--JB: Ben, this is where I'd like some of the details about min, max, etc. What proportion of people are on the wiki for more than a year, and how long do they stay on? -->

@@

## Methods of data collection and classification

@@

The entire R code [@rlanguage] and data files used for all the analyses and visualizations contained in this paper are openly available at https://doi.org/xxx/xx [ for peer reviewers the link is https://zenodo.org/records/14798942?preview=1&token=eyJhbGciOiJIUzUxMiJ9.eyJpZCI6ImMyYWE2MDc3LTMxMzAtNDM1Yi1hYTJkLWMyMGMyNWVhOWMyYiIsImRhdGEiOnt9LCJyYW5kb20iOiJiMjkyMTAyN2UwNzEwNWQ2OWIyZmQ4OTkwMjU0MzY1YyJ9.slnC1BbNgZkiJjrRqmKFK9a1bpI-e9VBbjcU7L5OEeYFs3whcL3POsfVj7phDnoaB5A_6mP9wpGirEaT9Dtvug] to enable re-use of materials and improve reproducibility and transparency [@Marwick2017]. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and the figures as CC-BY, to enable maximum re-use.

@@

Panel A of @fig-show-map-of-hiring-institution-and-cc shows the frequencies of hiring institutions according to their Carnegie Classification, a framework for classifying US colleges and universities according to the types of degrees awarded, levels of research activity, and topical foci [@shulman2001carnegie]. Doctoral universities with high and very high research activity are by far the most active institutions hiring archaeology faculty. Associate's colleges, also known as community colleges, rarely post job ads for archaeology faculty.

@@

Demand for jobs focusing on the Americas has been generally high over the past decade, with a peak in 2019-2020 and a subsequent decrease. Demand for jobs focusing on Africa was very low until 2019-2020, peaking in 2020-2021. The proportion of ads with a geographic focus on the Mediterranean has varied substantially, peaking in 2016 and experiencing a nadir in 2019, showing an inverse pattern to the Americas. Asia and India, the Near East, and Europe <!--crying face emoji BM: sorry! --> rarely occur as a geographical focus in job ads at US institutions. Asia and India, Africa, and the Americas appear correlated with each other, while the Near East and Mediterranean are inversely correlated in an opposite trend.

@@

Landscape archaeology, encompassing GIS and remote sensing, has remained prominent compared to other methods [@fig-show-metho-trends]. The popularity of this suite of skills may reflect its demand by employers in the Cultural Research Management (CRM) sector. @morganReadyNotArchaeological2023 found that almost one-third of 599 jobs ads posted by CRM employers sought candidates with experience in GIS. Methods focused on a specific element of the archaeological record, such as Lithic analysis, Zooarchaeology and Ceramics are among the least frequently mentioned in job ads. Instead, more popular methods are those that are relevant to multiple elements of the archaeological record (e.g. Archaeobotany encompasses macroscopic and microscopic plant remains; Bioarchaeology often includes skeletal analysis, isotopes, proteins, etc.).

@@

This discrepancy in our data set—wherein advertisements for tenure-track positions dominate the wiki despite their increasing rarity in larger market—may be due to several factors. The first involves the more limited circulation of advertising for short-term positions relative to advertising for tenure-track jobs. Many of these short-term positions are not advertised nationally, but are instead disseminated through local email lists and are filled by people close to the hiring department, such as recently graduated students. A second factor is bias in our data—with most Wiki users likely seeking a tenure-track job, non-tenure-track jobs may have been less frequently added to the Academic Jobs Wiki because they were peripheral to the goal of most users. This bias limits the reliability of our results on the ratio of tenure-track to non-tenure-track positions. Tellingly, Passalacqua’s examination of a similar wiki over a seven-year span showed a similar dominance of permanent positions—291/474 (63%) were for tenure-track hires at various levels (2018: Table 2). This suggests that these patterns are a reflection of the wiki, rather than the market itself. <!--JB MAYBE ADD A NOTE HERE FROM A CRISIS PAPER, PERHAPS MELISSA'S, ABOUT DISPARITY IN TRAINING AND APPLIED ANTHROPOLOGY. BM: [inserting earplugs] GO FOR IT! -->

@@

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. Over the course of the study period, we document a gradual decline in topics such as the Archaeology of complex societies and Archaeological science, the geographic foci of Mesoamerica and South America, and methods relating to Landscape archaeology <!--There is some shifting throughout the paper as to whether we capitalize the first letter of region/methods/topics or not. I think we should be consistent and capitalize throughout-->. These trends are harder to explain as we cannot link their origins to a historical event such as the COVID-19 pandemic. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, relational perspectives, in combination with human subjectivity and cultural framing, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism [@hodderArchaeologicalProcessIntroduction1999; @johnson2019archaeological; @Fogelin2019] <!--Should we cite Cate Frieman here? BM: grimace emoji -->. This theoretical trend might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical laboratory methods for technological, functional, and compositional analyses. Other factors relevant to this decline may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past and an increased concern with climate change, environmental sustainability, and resilience, shifting attention away from the study of monumental architecture, elite societies, political hierarchies, and state systems.

@@

Thanks to the anonymous contributors to the Academic Jobs Wiki, without whose efforts to collect and organise hundreds of job ads, this paper would not exist. To those of you who contributed to the wiki but did not obtain a tenure-track job, and/or decided to pivot to another industry, we hope this paper shows your time on the wiki was not wasted. Your efforts have helped contribute to more accurate expectation-setting for future generations of archaeologists.
